A Practical Optional Type System for Clojure

### 2012-10-28 dissertation/intro.tex
Improve Occurrence typing example

It is common in Clojure, like other dynamically-typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2} conditional tests are used
to refine the type of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing\cite{TF10} is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in section \ref{sec:OccurrenceTyping}).

\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure, label=lst:numvec2]
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [a b]
  [(if a a 0) 
   (if b b 0)])
\end{lstlisting}

To check this example, occurrence typing infers type information based on the result of each test.
In Clojure, \lstinline|nil| and \lstinline|false| are false values and all other values are true.
The test \lstinline|(if a a 0)| follows the \emph{then} branch is \lstinline|a| is not \lstinline|nil|
or \lstinline|false|. When checking this branch, we can safely refine the type of \lstinline|a| to \lstinline|Number| from
\lstinline|(U nil Number)|. Similarly, following the \emph{else} branch refines the type of \lstinline|a|
to \lstinline|nil| from \lstinline|(U nil Number)|.

>>>>

It is common in Clojure, like other dynamically-typed languages, to
encode implicit type invariants in conditional tests.
In listing \ref{lst:numvec2}, conditional tests are used
to refine the types of the bindings \lstinline|a| and \lstinline|b|.
Occurrence typing~\cite{TF10} is a technique useful for capturing these kinds
of type invariants (occurrence typing is discussed in further detail in Section \ref{sec:OccurrenceTyping}).

\begin{lstlisting}[caption=Example of occurrence typing in Typed Clojure, label=lst:numvec2]
(ann num-vec2 
     [(U nil Number) (U nil Number) -> (Vector* Number Number)])
(defn num-vec2 [x y]
  [(if x x 0) 
   (if y y 0)])
\end{lstlisting}

To check Listing \ref{lst:numvec2}, occurrence typing infers type information based on the results of conditional tests.
A conditional in Clojure is written \lstinline|(if test then else)|, where \lstinline|then|
is executed if \lstinline|test| is a true value, or \lstinline|else| is executed if \lstinline|test|
is a false value.
In Clojure, \lstinline|nil| and \lstinline|false| are the only false values: all other values are true values.
\lstinline|(if x x 0)| in Listing \ref{lst:numvec2} chooses a branch based on the truth value of the test expression \lstinline|x|.
Before this expression, the binding \lstinline|x| is known to be of type \lstinline|(U nil Number)|.
There are two cases to consider when checking this conditional: when the test \lstinline|x| is a true value, and when it is a false value.
If \lstinline|x| is a true value, then the binding \lstinline|x| must be of type \lstinline|Number|.
If \lstinline|x| is a false value, then the binding \lstinline|x| must be of type \lstinline|nil|.
Occurrence typing understands these invariants and applies them when checking each branch:
occurrences of \lstinline|x| in the first branch are considered to be of type \lstinline|Number|, and
occurrences of \lstinline|x| in the second branch are considered to be of type \lstinline|nil|.
This infers \lstinline|[(if x x 0) (if y y 0)])| to be of type
\lstinline|(Vector* Number Number)| (the type for a two-place heterogeneous vector of \lstinline|Number|s).
